Global land characterisation using land cover fractions at 100 m resolution

### 2021-02-04 article/article.tex
Move parts to appendix

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{RRMSE}{RRMSE}{relative root mean squared error}
\newacronym{RMAE}{RMAE}{relative mean absolute error}
\newacronym{RME}{RME}{relative mean error}
\newacronym{NSE}{NSE}{the Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{the International Institute for Applied Systems Analysis}
\newacronym{WUR}{WUR}{Wageningen University \& Research}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NDMI}{NDMI}{Normalised Difference Moisture Index}
\newacronym{NIRv}{NIR\textsubscript{v}}{\gls{NIR} of vegetation}
\newacronym{EVI}{EVI}{Enhanced Vegetation Index}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SIF}{SIF}{sun-induced chlorophyll fluorescence}
\newacronym{GPP}{GPP}{gross primary productivity}
\newacronym{MSI}{MSI}{multi-spectral instrument}
\newacronym{SAR}{SAR}{synthetic aperture radar}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{classification and regression tree}
\newacronym{GLM}{GLM}{general linear regression model}
\newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
\newacronym{GDAL}{GDAL}{the Geospatial Data Abstraction Library}
\newacronym{GFCC}{GFCC}{Global Forest Cover Change}
\newacronym{GHSL}{GHSL}{Global Human Settlement Layer}
\newacronym{GSW}{GSW}{Global Surface Water}
%\makenoidxglossaries

@@

We compared the performance of multiple types of linear and machine learning regression algorithms for land cover fraction mapping.
For the model with the lowest MAE, we assessed the importance of six categories of features, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe.
The models were trained on over 138\,000 points and validated on a separate dataset of over 20\,000 points, provided by the \gls{CGLS-LC100} project. Both datasets are global and aligned with the PROBA-V 100 m UTM grid.

A major challenge for land cover fraction mapping models is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% fractions are difficult to predict accurately. We proposed a new solution by combining three models. A binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.

Results showed that from the compared models, random forest regression reached the lowest RMSE of 17.3\%. Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved using random forest with our proposed three-model approach. This model was driven primarily by optical remote sensing data, followed by climate data.

This research proves that machine learning algorithms can be applied globally to map a wide variety of land cover fractions. Fraction mapping expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which enables customising the result for a diverse range of uses. The three-step approach enables mapping 0\% and 100\% fractions more accurately, and thus has already been taken up in the operational production of global land cover fraction layers within the \gls{CGLS-LC100} project. This study also includes an accuracy assessment of land cover fraction maps both thematically and spatially, which could be taken up by future land cover fraction mapping efforts.
\end{abstract}

@@

In this study we investigated and compared a variety of methods for performing global land cover fraction mapping.
Furthermore, we proposed a method to enhance land cover fraction mapping further by dealing with data imbalance, which is inherent in land cover fraction training data, and tested it on the method that achieved the highest accuracy in the method comparison.
Lastly, we investigated the effect of a wide variety of input features on each mapped class fraction, to gain insight into what data was key for the model with the highest accuracy to achieve the best result.
Therefore the objectives of our study were to:

\begin{enumerate}
 \item Compare the performance of a variety of machine learning regression models for mapping fractions of land cover on a global scale.
 \item Investigate approaches for reducing bias in the model predictions with regards to zero inflation and predictions tending towards the mean.
 \item Analyse the importance of features input into the highest accuracy model for the prediction of each land cover fraction.
 \item Assess the accuracy of the models from a thematic and spatial point of view, comparing it to existing global land cover products.
\end{enumerate}

@@

To train the models and predict land cover fractions in unsampled locations, six groups of features were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see supplementary material, table S1).
These features had to be preprocessed before they could be input into the models.

@@

We masked out the clouds from the time series of each of the four PROBA-V spectral bands, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were further than 2 standard deviations away from the \ac{LOESS} curve fitted over the blue reflectance band.
We then used the resulting cloud-free time series to generate the following \glspl{VI}: \gls{NDVI}, \gls{NDMI}, \gls{EVI}, \gls{OSAVI} and \gls{NIRv}.
These \glspl{VI} are commonly used in land cover mapping to aid in discerning the vegetation classes among each other.
Next, we used the \gls{VI} time series to calculate the descriptive statistics (median and interquartile range) over the whole time series, as well as for each phenological season separately.
We included the resulting metrics as vegetation index features.
Next, we ran harmonic analysis \citep{jakubauskas2001harmonic} on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
The temporal metrics from the harmonic model quantify the seasonality of the area and allow differentiating between vegetation with different seasonality, such as crops.
From this harmonic analysis we obtained the minimum and maximum values of the \gls{NDVI} time series, which we included as additional vegetation index features.
We also obtained the the trend and intercept components from the harmonic model.
Lastly, we calculated the phase and amplitude for the two harmonic orders from the respective sine and cosine components of the model.
We used these sine, cosine, trend and intercept components, as well as the phase and amplitude of the two harmonic orders, as temporal features in our study.
For a list of features that were ultimately used as input to the models, see supplementary material, table S1.

@@

After the feature selection process, 67 features remained.
These features include data from each of the feature categories.
See supplementary material, table S1 for an overview of all of the features that remained and thus used in model training and prediction.

@@

We compared a wide array of machine learning regression methods for deriving land cover fractions.
The tested methods can be broadly divided into three types: linear models, machine learning models based on \glspl{CART}, and machine learning models not based on \glspl{CART}.
See table \ref{tab-methods} for the full list of methods that we compared in this study.
In addition, as a baseline we also compared the results with a trivial equal proportion model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).
We tuned each algorithm to select optimal parameters, and postprocessed the output of each algorithm as necessary to ensure that all land cover fractions in each pixel add up to 100\%.
Namely, if the model output for any class was outside of the 0-100\% range, the values were clamped to that range, and if the values did not add up to 100\%, they were linearly rescaled so that they would.
All of the model building and data analysis was performed using the free and open-source statistical software R \citep{r_2019}.

\begin{table}
\centering
    \begin{tabular}{m{4cm}ll}
         \toprule
         \textbf{Category} & \textbf{Name} & \textbf{Reference} \tabularnewline
         \midrule
         Linear models &
         \begin{tabular}{@{}l@{}}
         \Glsentryfull{FNC}\tabularnewline
         \Glsentryfull{GLM}\tabularnewline
         \Glsentryfull{PLS} regression \tabularnewline
         Lasso regression \tabularnewline
         \Glsentryfull{MLR}
         \end{tabular}
         & 
         \begin{tabular}{@{}l@{}}
         \citealt{keller_fuzzy_1985} \tabularnewline
         \citealt{neter_applied_1996} \tabularnewline
         \citealt{wold_pls-regression_2001} \tabularnewline
         \citealt{tibshirani_regression_1996} \tabularnewline
         \citealt{theil_multinomial_1969}
         \end{tabular}
         \tabularnewline \midrule
         Machine learning models based on decision trees &
         \begin{tabular}{@{}l@{}}
         \Glsentryfull{RF} regression\tabularnewline
         Cubist regression
         \end{tabular}
         &
         \begin{tabular}{@{}l@{}}
         \citealt{breiman2001random} \tabularnewline
         \citealt{cubist}
         \end{tabular}
         \tabularnewline \midrule
         Other machine learning models &
         \begin{tabular}{@{}l@{}}
         \Gls{MLP} \glsentryfullpl{NN}\tabularnewline
         \Glsentryfull{SVM} regression
         \end{tabular}
         & 
         \begin{tabular}{@{}l@{}}
         \citealt{dreyfus_artificial_1990} \tabularnewline
         \citealt{suykens_least_1999}
         \end{tabular}
         \tabularnewline \midrule
         Ensemble learning &
         Super Learner
         & 
         \citealt{laan_super_2007}
         \tabularnewline \bottomrule
    \end{tabular}
    \caption{List of regression methods for land cover fraction estimation tested in this study.}
    \label{tab-methods}
\end{table}

\subsubsection{Linear models}

We chose to compare five types of linear models, to have a baseline for a comparison with the nonlinear machine learning models.
The most simple model we selected was the \gls{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression.
It is an extension to the standard linear regression that allows for multiple outcomes, and is implemented in base R \citep{r_2019}.
Next, we tested two linear models that include input data regularisation in the model itself: lasso regression \citep{tibshirani_regression_1996}, implemented in the \cran{glmnet} package \citep{glmnet}, and \gls{PLS} regression \citep{wold_pls-regression_2001}, implemented in the \cran{pls} package \citep{pls}.
We also tested \gls{MLR} \citep{theil_multinomial_1969}, provided by the \cran{nnet} package \citep{nnet}.
\gls{MLR} is usually used for classification and is fit using land cover class labels rather than fractions, but the output includes probabilities for each class that add up to 100\%.
We fit \gls{MLR} using the dominant land cover class as a label for the pixel and used the class probabilities as a proxy for land cover fractions.
Lastly, we tested \gls{FNC} regression \citep{keller_fuzzy_1985}, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, which is provided in the \cran{GSIF} package \citep{hengl2004fuzzycmeans}.
It is a simple regression method, where the land cover fractions in a pixel are determined by the distance of the pixel from the centroids of each class in feature space.

\subsubsection{Non-tree machine learning methods}

\Glsentryfullpl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\% with no need for additional postprocessing.
In this study, after performing tuning, we ended up using a \glsentryfull{MLP} \citep{dreyfus_artificial_1990} with three hidden layers with 128, 64 and 32 neurons per layer respectively.
We used the Nadam optimiser \citep{dozat_incorporating_2016} with \gls{MAE} as the loss function to optimise the \gls{NN}, and softmax activation for the output.
The models were trained using the \cran{keras} package \citep{keras}, built upon TensorFlow that enables the use of a graphics processing unit to accelerate the \gls{NN} training process.

\Glsentryfullpl{SVM} are machine learning models that attempt to find the optimal boundary between the class clusters in feature space by constructing a dividing hyperplane.
For land cover fraction classification, we used \gls{SVM} regression based on Least Squares \gls{SVM} \citep{suykens_least_1999}.
As \gls{SVM} models are univariate, we used the binary relevance method \citep{karalas2016br}: training separate models per class that predict a single class, and then combining the results.
We found the package \cran{liquidSVM} \citep{liquidSVM} suitable for this purpose, as it is optimised for processing large datasets such as ours.

\subsubsection{Tree-based machine learning methods}

\Glsentryfull{RF} regression is a popular method for land cover classification that works by building a number of \gls{CART} decision trees based on random subsets of the input training data, and taking the mean or median of the ``votes'' of these individual decision trees \citep{breiman2001random}.
\ac{RF} is univariate, therefore we again used the binary relevance approach.
The processing was done using the \cran{ranger} package \citep{ranger}.

Lastly, we tested Cubist regression \citep{cubist}.
It is based on \gls{RF} regression, but instead of using a threshold of a feature to split the decision tree, Cubist uses a linear regression based on a subset of the data relevant for the split in question.
In addition, it features committees, a boosting technique that iteratively trains trees so as to learn from the previously generated ones.
Model tuning led to us using 10 committees.
Cubist predictions were also made using the binary relevance method.
The algorithm is provided by the \cran{Cubist} package \citep{cubistpackage}.

\subsubsection{Ensemble learning}

Lastly, we made an ensemble from the two machine learning models that produced the lowest \gls{RMSE} and \gls{MAE} respectively.
We used the \cran{sl3} package \citep{sl3} to create a hierarchical ensemble, where the two models were cross-validated using 10-fold cross-validation to obtain relative weights of each model for each land cover class, following the super learner algorithm \citep{laan_super_2007}.
The predictions of the two models, along with the weights of the models, were then used as input features for another \gls{RF} regression metalearner.
The output from this metalearner is the final prediction of the whole ensemble.
As the ensembled methods are univariate, we used the binary relevance method in this case as well.

%Another model that was considered was \cran{SuperLearner} \citep{SuperLearner}, an ensembling method that can combine multiple of the tested methods into one model, giving models different weights according to how well they are known to perform for a given task.
%Here we ran into similar issues, as cross-validation is required for each model to train, which increases processing time dramatically.
%Attempts to reduce training time by lowering the number of folds resulted in worse results than when using a single classifier.

@@

To assess the performance of the models, we used a number of statistical measures.
We started with the statistics of assessing land cover fraction model accuracy that are the most commonly used in this field: \glsentryfull{RMSE}, \glsentryfull{MAE} and \glsentryfull{ME}.
\gls{MAE} represents the average difference between the predicted and the reference land cover fractions.
In our case, its unit is percentage points.
\gls{RMSE} squares the errors, therefore giving a larger penalty for large errors, and thus is always higher than \gls{MAE}.
These statistics are relatively straightforward to calculate and interpret.
In the case of land cover fraction mapping, \gls{RMSE} is very sensitive to errors where a pixel is entirely mapped as a different class (i.e. 100\% instead of 0\%).
\gls{MAE} is more lenient and not as influenced by a small number of such large misclassifications.
Thus it is more indicative of the overall model accuracy, whereas \gls{RMSE} is more indicative of the presence of large errors.

@@

Next, we estimated the goodness of fit of the models by calculating the coefficient of determination $R^2$ of the models in two ways.
The first way is \gls{NSE} \citep{nash1970river}, which is equivalent to an $R^2$ of a linear regression model whose intercept and slope terms are predetermined and are equal to 0 and 1, respectively.
This metric shows how far away the predicted values are from the 1:1 line with the reference values.
The value range of \gls{NSE} is $(-\infty, 1]$, where a value of 0 means that the predicted values are no better than predicting the mean value.
The second way is to calculate the $R^2$ of an Ordinary Least Squares regression that estimates the intercept and slope, rather than presetting it.
This method always gives higher $R^2$ values than \gls{NSE}, as it allows the regression line to have more flexibility.
The two ways of estimating the coefficient of determination are herewith represented as $R^2_{NSE}$ and $R^2_{OLS}$, respectively.

@@

Moreover, to show the variation in predicted land cover fraction values depending on the magnitude of the fractions, we produced boxplots showing the distribution of the predictions.
The boxplots are binned for each 10\% of the predictions.
The 1:1 line indicates how well do the distributions of the predictions and the reference data match.

@@

\subsection{Feature importance}

@@

\subsection{Method accuracy comparison}

The overall accuracy statistics of the compared models are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
Overall, \gls{RF} regression achieved the highest accuracy: by \gls{RMSE} when using a single model and a mean vote, and by \gls{MAE} when using a median vote.
The combination of the three-step model and the median vote resulted in a decrease in \gls{RMSE} without impacting \gls{MAE}.

\begin{table}
\centering
%\adjustbox{max width=0.5\textwidth}{
\begin{tabular}{lllllll}
\toprule
\textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & $\mathbf{R^2_{NSE}}$ & $\mathbf{R^2_{OLS}}$ & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
\midrule
Equal proportions
& 29.9  & 21.4  & 0     & — & 26±5  & 0.13±0.07 \\
\Gls{FNC}
& 24.4  & 13.5  & 0.33  & 0.35 & 53±4  & 0.42±0.06 \\
\Gls{GLM}, \Gls{PLS}, Lasso
& 21.6  & 12.7  & 0.48  & 0.49 & 56±4  & 0.43±0.05 \\
\Gls{MLR}
& 21.6  & 12.1  & 0.48  & 0.48 & 58±4  & 0.46±0.06 \\
\Gls{MLP} \glspl{NN}
& 22.7  & 9.2   & 0.43  & 0.52 & 68±1  & 0.57±0.02 \\
\Gls{SVM} regression
& 20.7  & 8.9   & 0.52  & 0.56 & 69±2  & 0.58±0.03 \\
Cubist regression
& 18.1  & 8.1   & 0.63  & 0.65 & \textbf{72±2}  & 0.63±0.03 \\
\Gls{RF} regression
& \textbf{17.3}  & 9.4   & \textbf{0.66}  & \textbf{0.67} & 67±4  & 0.57±0.05 \\
\ensuremath{''}
only RS features
& 18.4  & 10.3  & 0.62  & 0.64 & 64±4  & 0.54±0.05 \\
\ensuremath{''} two-step
& 19.9  & 8.2   & 0.56  & 0.60 & 71±2  & 0.62±0.02 \\
\ensuremath{''} three-step
& 19.4  & 8.4   & 0.58  & 0.60 & 71±3  & 0.62±0.04 \\
\ensuremath{''} median vote
& 20.7  & \textbf{7.9}   & 0.52  & 0.60 & 71±1  & 0.63±0.02 \\
\ensuremath{''} \ensuremath{''} two-step
& 20.0  & 8.1   & 0.54  & 0.60 & \textbf{72±1}  & 0.63±0.02 \\
\ensuremath{''} \ensuremath{''} three-step
& 20.2  & \textbf{7.9}   & 0.54  & 0.60 & \textbf{72±2}  & \textbf{0.64±0.02} \\
Cubist + RF ensemble
& 17.7  & 8.6   & 0.65  & 0.65 & 70±3 & 0.61±0.04 \\
\bottomrule
\end{tabular}
%}
\caption{Accuracy statistics of the tested models. Best performing statistics are highlighted. ``Only RS features'' stands for a model trained only with \glspl{VI} and temporal metrics.}
\label{tab-accuracy}
\end{table}

All tested models had lower accuracy when estimating the cover fractions of vegetation classes (herbaceous, trees, crops, shrubs), compared to the non-vegetated land cover (inland water, built-up, bare land), as shown in figure \ref{fig-models}.
This was exacerbated by the imbalance in class distribution: the built-up class is rare and very rarely forms a majority, therefore a prediction of 0\% leads to a perfect prediction most of the times, and even if it does not, the error is low.
In contrast, tree cover had a lot more balanced distribution of fractions, including a large number of pure pixels, which leaves no such trivial solution.
The \gls{RRMSE} and \gls{RMAE} statistics show this (see supplementary material, figure S1): the tree cover prediction has low relative error given its high mean value (32\%), whereas the built-up class is the most challenging to predict according to \gls{RRMSE} given its low mean value (3\%).
The shrub cover fraction is also very challenging to predict, since it had the highest \gls{RMAE} and none of the models showed large improvements in \gls{RMSE} compared to the equal proportions model.
There were only moderate improvements for predicting the herbaceous vegetation fraction as well.
In contrast, most models were a significant improvement in predicting tree, built-up, water and bare cover fractions compared to the equal proportion model, indicating that the features used to train the models were useful to distinguish these classes from the others and to quantify their proportions in each pixel.

@@

\subsection{Feature importance}

@@

% NOTE: Appendix is in supplement.tex

@@

%\listoffigures

>>>>

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{OLS}{OLS}{ordinary least squares}
\newacronym{RRMSE}{RRMSE}{relative root mean squared error}
\newacronym{RMAE}{RMAE}{relative mean absolute error}
\newacronym{RME}{RME}{relative mean error}
\newacronym{NSE}{NSE}{the Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{the International Institute for Applied Systems Analysis}
\newacronym{WUR}{WUR}{Wageningen University \& Research}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NDMI}{NDMI}{Normalised Difference Moisture Index}
\newacronym{NIRv}{NIR\textsubscript{v}}{\gls{NIR} of vegetation}
\newacronym{EVI}{EVI}{Enhanced Vegetation Index}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SIF}{SIF}{sun-induced chlorophyll fluorescence}
\newacronym{GPP}{GPP}{gross primary productivity}
\newacronym{MSI}{MSI}{multi-spectral instrument}
\newacronym{SAR}{SAR}{synthetic aperture radar}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{classification and regression tree}
\newacronym{GLM}{GLM}{general linear regression model}
\newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
\newacronym{GDAL}{GDAL}{the Geospatial Data Abstraction Library}
\newacronym{GFCC}{GFCC}{Global Forest Cover Change}
\newacronym{GHSL}{GHSL}{Global Human Settlement Layer}
\newacronym{GSW}{GSW}{Global Surface Water}
%\makenoidxglossaries

@@

A major challenge for land cover fraction mapping models is data sparsity: the dataset is zero-inflated due to how common the 0\% fraction is, but as regression favours the mean, 0\% and 100\% fractions are difficult to predict accurately. We proposed a new solution by combining three models: a binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.

We compared the performance of multiple types of linear and machine learning regression algorithms for land cover fraction mapping, and for our proposed three-step model, assessed the importance of six categories of features, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe.
The models were trained on over 138\,000 points and validated on a separate dataset of over 20\,000 points, provided by the \gls{CGLS-LC100} project. Both datasets are global and aligned with the PROBA-V 100 m UTM grid.

Results showed that the random forest regression model reached the lowest RMSE of 17.3\%. Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved using random forest regression with our proposed three-model approach and median vote. This model was driven primarily by optical remote sensing data.

This research proves that machine learning algorithms can be applied globally to map a wide variety of land cover fractions, and tuned for user needs. Fraction mapping expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which enables customising the result for a diverse range of uses. The three-step approach enables mapping 0\% and 100\% fractions more accurately, and thus has already been taken up in the operational production of global land cover fraction layers within the \gls{CGLS-LC100} project. This study also includes an accuracy assessment of land cover fraction maps both thematically and spatially, which could be taken up by future land cover fraction mapping efforts.
\end{abstract}

@@

In this study, after investigating and comparing a variety of methods for performing global land cover fraction mapping, we proposed a method to enhance land cover fraction mapping by dealing with the inherent data imbalance issue.
We assessed the accuracy of the result both thematically and spatially, by calculating model accuracy spatially across the globe.
Lastly, we investigated the effect of input features on each mapped class fraction.
Therefore the primary objectives of our study were to:

\begin{enumerate}
 \item Investigate approaches for reducing bias in the model predictions with regards to zero inflation and predictions tending towards the mean.
 \item Assess the accuracy of the models from a thematic and spatial point of view, comparing it to existing global land cover products.
\end{enumerate}

@@

To train the models and predict land cover fractions in unsampled locations, six groups of features were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see \ref{sec-inputdata}, table \ref{tab-inputdata}).
These features had to be preprocessed before they could be input into the models.

@@

We masked out the clouds from the time series of each of the four PROBA-V spectral bands, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were further than 2 standard deviations away from the \ac{LOESS} curve fitted over the blue reflectance band.
We then used the resulting cloud-free time series to generate the following \glspl{VI}: \gls{NDVI}, \gls{NDMI}, \gls{EVI}, \gls{OSAVI} and \gls{NIRv}.
These \glspl{VI} are commonly used in land cover mapping to aid in discerning the vegetation classes among each other.
Next, we used the \gls{VI} time series to calculate the descriptive statistics (median and interquartile range) over the whole time series, as well as for each phenological season separately.
We included the resulting metrics as vegetation index features.
Next, we ran harmonic analysis \citep{jakubauskas2001harmonic} on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
The temporal metrics from the harmonic model quantify the seasonality of the area and allow differentiating between vegetation with different seasonality, such as crops.
From this harmonic analysis we obtained the minimum and maximum values of the \gls{NDVI} time series, which we included as additional vegetation index features.
We also obtained the the trend and intercept components from the harmonic model.
Lastly, we calculated the phase and amplitude for the two harmonic orders from the respective sine and cosine components of the model.
We used these sine, cosine, trend and intercept components, as well as the phase and amplitude of the two harmonic orders, as temporal features in our study.
For a list of features that were ultimately used as input to the models, see \ref{sec-inputdata}, table \ref{tab-inputdata}.

@@

After the feature selection process, 67 features remained.
These features include data from each of the feature categories.
See \ref{sec-inputdata}, table \ref{tab-inputdata} for an overview of all of the features that remained and thus used in model training and prediction.

@@

We compared a wide array of machine learning regression methods for deriving land cover fractions.
The tested methods can be broadly divided into three types: linear models, machine learning models based on \glspl{CART}, and machine learning models not based on \glspl{CART}.
See table \ref{tab-methods} for the full list of methods that we compared in this study.
In addition, as a baseline we also compared the results with a trivial equal proportion model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).
Any useful model has to perform better than the equal proportion model, and by comparison to it, it is possible to tell how much better a model performs in estimating the fraction of each class.
We tuned each algorithm to select optimal parameters, and postprocessed the output of each algorithm as necessary to ensure that all land cover fractions in each pixel add up to 100\%.
Namely, if the model output for any class was outside of the 0-100\% range, the values were clamped to that range, and if the values did not add up to 100\%, they were linearly rescaled so that they would.
All of the model building and data analysis was performed using the free and open-source statistical software R \citep{r_2019}.
The results showed that \gls{RF} regression performed the best, with the lowest \gls{RMSE} value of 17.3\%.
For more technical details and in-depth results of the regression model accuracy comparison, see \ref{sec-model-comparison}.

\begin{table}
\centering
  \begin{adjustbox}{max width=\textwidth}
    \begin{tabular}{m{4cm}lll}
         \toprule
         \textbf{Category} & \textbf{Name} & \textbf{Reference} & \textbf{R package and authors} \tabularnewline
         \midrule
         Linear models &
         \begin{tabular}{@{}l@{}}
         \Glsentryfull{FNC}\tabularnewline
         \Glsentryfull{GLM}\tabularnewline
         \Glsentryfull{PLS} regression \tabularnewline
         Lasso regression \tabularnewline
         \Glsentryfull{MLR}
         \end{tabular}
         & 
         \begin{tabular}{@{}l@{}}
         \citealt{keller_fuzzy_1985} \tabularnewline
         \citealt{neter_applied_1996} \tabularnewline
         \citealt{wold_pls-regression_2001} \tabularnewline
         \citealt{tibshirani_regression_1996} \tabularnewline
         \citealt{theil_multinomial_1969}
         \end{tabular}
         &
         \begin{tabular}{@{}l@{}}
         \cran{GSIF} \citep{hengl2004fuzzycmeans} \tabularnewline
         stats \citep{r_2019} \tabularnewline
         \cran{pls} \citep{pls} \tabularnewline
         \cran{glmnet} \citep{glmnet} \tabularnewline
         \cran{nnet} \citep{nnet}
         \end{tabular}
         \tabularnewline \midrule
         Machine learning models based on decision trees &
         \begin{tabular}{@{}l@{}}
         \Glsentryfull{RF} regression\tabularnewline
         Cubist regression
         \end{tabular}
         &
         \begin{tabular}{@{}l@{}}
         \citealt{breiman2001random} \tabularnewline
         \citealt{cubist}
         \end{tabular}
         &
         \begin{tabular}{@{}l@{}}
         \cran{ranger} \citep{ranger} \tabularnewline
         \cran{Cubist} \citep{cubistpackage}
         \end{tabular}
         \tabularnewline \midrule
         Other machine learning models &
         \begin{tabular}{@{}l@{}}
         \Gls{MLP} \glsentryfullpl{NN}\tabularnewline
         \Glsentryfull{SVM} regression
         \end{tabular}
         & 
         \begin{tabular}{@{}l@{}}
         \citealt{dreyfus_artificial_1990} \tabularnewline
         \citealt{suykens_least_1999}
         \end{tabular}
         & 
         \begin{tabular}{@{}l@{}}
         \cran{keras} \citep{keras} \tabularnewline
         \cran{liquidSVM} \citep{liquidSVM}
         \end{tabular}
         \tabularnewline \midrule
         Ensemble learning &
         Super Learner
         & 
         \citealt{laan_super_2007}
         &
         \cran{sl3} \citep{sl3}
         \tabularnewline \bottomrule
    \end{tabular}
  \end{adjustbox}
    \caption{List of regression methods for land cover fraction estimation tested in this study.}
    \label{tab-methods}
\end{table}

@@

To assess the performance of the models, we used a number of statistical measures.
We started with the statistics of assessing land cover fraction model accuracy that are the most commonly used in this field: \glsentryfull{RMSE}, \gls{MAE} and \gls{ME}.
\gls{MAE} represents the average difference between the predicted and the reference land cover fractions.
In our case, its unit is percentage points.
\gls{RMSE} squares the errors, therefore giving a larger penalty for large errors, and thus is always higher than \gls{MAE}.
These statistics are relatively straightforward to calculate and interpret.
In the case of land cover fraction mapping, \gls{RMSE} is very sensitive to errors where a pixel is entirely mapped as a different class (i.e. 100\% instead of 0\%).
\gls{MAE} is more lenient and not as influenced by a small number of such large misclassifications.
Thus it is more indicative of the overall model accuracy, whereas \gls{RMSE} is more indicative of the presence of large errors.

@@

Next, we estimated the goodness of fit of the models by calculating the coefficient of determination $R^2$ of the models in two ways.
The first way is \gls{NSE} \citep{nash1970river}, which is equivalent to an $R^2$ of a linear regression model whose intercept and slope terms are predetermined and are equal to 0 and 1, respectively.
This metric shows how far away the predicted values are from the 1:1 line with the reference values.
The value range of \gls{NSE} is $(-\infty, 1]$, where a value of 0 means that the predicted values are no better than predicting the mean value.
The second way is to calculate the $R^2$ of an \gls{OLS} regression that estimates the intercept and slope, rather than presetting it.
This method always gives higher $R^2$ values than \gls{NSE}, as it allows the regression line to have more flexibility.
The two ways of estimating the coefficient of determination are herewith represented as $R^2_{NSE}$ and $R^2_{OLS}$, respectively.

@@

Moreover, to show the variation in predicted land cover fraction values depending on the magnitude of the fractions, we produced boxplots showing the distribution of the predictions.
The boxplots are binned for each 10\% of the predictions.
The 1:1 line indicates how well do the distributions of the predictions and the reference data match.
We also created additional plots showing how \gls{RMSE}, \gls{MAE} and \gls{ME} change over these bins.

@@

Lastly, we investigated how features affect model accuracy. See \ref{sec-varimp} for more information on feature importance.

@@

\subsection{Regression model accuracy comparison}

Overall, \gls{RF} regression achieved the highest accuracy: by \gls{RMSE} when using a single model and a mean vote, and by \gls{MAE} when using a median vote.
For more details about the results of the accuracy comparison between the tested regression models, see \ref{sec-model-comparison}.

@@

@@

@@

\appendix

\section{Regression model comparison} \label{sec-model-comparison}

\subsection{Regression models tested in this study}
